Supplemental Materials for ``Universals of word order result from optimization of grammars for efficient communication''

### 2019-01-25 second-draft/paper-names.tex
attempting revision

\begin{abstract}
The universal properties of human languages have been the subject of intense study across disciplines. We report novel computational and corpus-based evidence for the hypothesis that a prominent subset of these universal properties---those related to word order---result from a process of optimization for efficient communication among humans. We develop a probabilistic, differentiable model of word order grammars: the means by which different languages convert underlying hierarchical structures into strings of words. We show how the parameters of a word order grammar can be optimized for robustness of information transfer, as quantified using mutual information with latent tree structures, and efficiency of processing, as quantified using incremental predictability. Applying these grammars to tree structures found in dependency corpora, we show that optimizing the grammar parameters for efficiency and robustness results in word order patterns that reproduce a large subset of the major word order correlations reported in the linguistic typological literature, and reproduce the predictions of previous heuristic theories such as dependency length minimization.
\end{abstract}

@@

\dropcap{F}or decades, researchers in fields ranging from philology to cognitive science to statistical physics have been involved in documenting and trying to explain the universal syntactic and statistical properties of human language \cite{behaghel1909beziehungen,zipf1949human,greenberg1963universals,lin2017critical}. An explanation for the universal properties of language would enable a deeper scientific understanding of what human language is and how to model it, with applications in psychology and natural language processing \cite{hawkins2007processing,bender2009linguistically,bender2013linguistic}. In this work we examine this question from a computational perspective, demonstrating a fully formalized and implemented framework in which certain syntactic universals can be explained through the statistical optimization of grammars for information-theoretic efficiency. % TODO add a cogsci person to citation chain at the beginning

We aim to explain the linguistic universals related to word order first documented by Greenberg \cite{greenberg1963universals}. These universals are called \key{implicational} universals, and take the form of correlations between certain word order patterns and others across languages. For example, whether a language puts verbs before or after objects strongly determines whether it places adjectives before or after nouns, and whether it had prepositions or postpositions \cite{dryer1992greenbergian}.

@@

Researchers working in the functional paradigm have proposed a range of criteria which languages should meet in order to enable efficient communication and processing. These criteria are based on theories of online processing difficulty (see \cite{jaeger2011language} for a review) and on information theoretic notions of efficiency and robustness in communication \cite{ferrericancho2001two,piantadosi2011word,gibson2013noisy}, and they have been formalized to varying degrees.

@@

We formalize the notion of communicative efficiency using two terms: one having to do with minimizing the difficulty of linguistic communication, and one having to do with maximizing the information transferred in such communication. We quantify the difficulty of linguistic communication in terms of the word-by-word \key{predictability} of strings, which has been found to be a highly accurate and general predictor of human online comprehension difficulty \cite{hale2001probabilistic,levy2008expectation,smith2013effect}, and which can be justified in terms of predictive coding theories of information processing in the brain \cite{friston2009predictive} as well as the general algorithmic complexity of language generation and comprehension \cite{li2008introduction}. We quantify the information transferred in language as \key{parseability}: the quantity of information contained in linguistic utterances about latent hierarchical tree structures. We also compare the results of optimizing for communicative efficiency to the results of optimizing for dependency length minimization, a previously proposed approximate metric of the amount of memory involved in language production and comprehension \cite{hudson1995measuring,gibson1998linguistic,gibson2000dependency,liu2017dependency,temperley2018minimizing}.

@@

We formalize predictability as follows. Given a language model $P_\phi$ with parameters $\phi$, the log probability of a sentence ${\bf w}$ is:
\begin{equation*}
R_{Pred}^\phi(t; {\bf w}) = \sum_{i=1}^{\#{\bf w}} \log P_\phi(w_{i+1}|w_{1\dots i}),
\end{equation*}
where the language model parameters $\phi$ are chosen predict the language induced by tree distribution $T$ and word order grammar $\L_\theta$ as accurately as possible (maximizing predictability):
\begin{equation*}
	\phi(\theta) = \argmax_\phi \displaystyle \E_{t \sim T} \E_{{\bf w} \sim L_\theta(t)} [ R^\phi_{Pred}({\bf w}, \phi) ].
\end{equation*}
We implement the language model $P_\phi$ using an LSTM neural network architecture \cite{hochreiter1997long}, currently the state of the art for this task \cite{goldberg2017neural}. Note that the language model parameters $\phi$ and the word order grammar parameters $\theta$ are recursively dependent on each other. See the SI for details on how $\theta$ and $\phi$ are jointly optimized, as well as on the neural network architecture.

@@

\subsection{Greenbergian Word Order Correlations}
\label{sec:greenberg}
\begin{table*}
	\begin{center}
\begin{tabular}{|ll|l|l|l|ll|l|}
	\hline
	%&
	\multicolumn{2}{|c|}{Correlates with...}   &          \multirow{2}{*}{Operationalization}       &     \multirow{2}{*}{Real}   & \multirow{2}{*}{DepL}  & \multirow{2}{*}{Pred}   &  \multirow{2}{*}{Pars} &  \multirow{2}{*}{Efficiency}  \\ 
	verb & object &      & & & &  & \\ \hline \hline % \textsc{verb} $\xrightarrow{obj}$ \textsc{noun}
%adp. &
adposition    &    NP    &    lifted\_case    &    86    &    \textbf{81}$^{***}$    &    47    &    \textbf{76}$^{***}$    &    \textbf{68}$^{***}$   \\
copula    &    NP    &    lifted\_cop    &    94    &    \textbf{81}$^{***}$    &    53    &    \textbf{79}$^{***}$    &    \textbf{61}$^{**}$   \\
auxiliary    &    VP    &    aux    &    88    &    \textbf{74}$^{***}$    &    \textbf{84}$^{***}$    &    55    &    \textbf{69}$^{**}$   \\
noun    &    genitive    &    nmod    &    80    &    \textbf{82}$^{***}$    &    55    &    \textbf{74}$^{***}$    &    \textbf{70}$^{***}$   \\
noun    &    relative clause    &    acl    &    80    &    \textbf{85}$^{***}$    &    48    &    \textbf{77}$^{**}$    &    \textbf{73}$^{***}$   \\
complementizer    &    S    &    lifted\_mark    &    76    &    \textbf{85}$^{***}$    &    \textbf{59}$^{**}$    &    \textbf{80}$^{***}$    &    \textbf{74}$^{**}$   \\
verb    &    PP    &    obl    &    88    &    \textbf{78}$^{***}$    &    \textbf{72}$^{***}$    &    59    &    \textbf{69}$^{**}$   \\
want    &    VP    &    xcomp    &    88    &    \textbf{90}$^{***}$    &    \textbf{78}$^{**}$    &    \textbf{92}$^{***}$    &    \textbf{92}$^{***}$   \\
verb    &    subject    &    nsubj    &    33    &    \textbf{29}$^{**}$    &    51    &    \textbf{8}$^{***}$    &    \textbf{13}$^{***}$   \\
verb    &    manner adverb    &    advmod    &    35    &    51    &    \textbf{21}$^{***}$    &    51    &    \textbf{32}$^{***}$   \\
 \hline
%comp. &
%\multirow{3}{*}{NP}&
%	AP &
    \multicolumn{6}{l}{\footnotesize{Significance levels: $^*$: $p < 0.05$, $^{**}$: $p < 0.01$, $^{***}$: $p < 0.001$}}
\end{tabular}
	\end{center}
	\caption{Greenbergian Correlations. Following \cite{dryer1992greenbergian}, each correlation is stated in terms of a pair of a `verb patterner' and an `object patterner', whose relative order correlate with that of verbs and objects.
%	For clarity, we organize the correlations by the category of the `verb patterner'.
	For each correlation, we give our operationalization in terms of UD. %A rightwards arrow indicates that the dependency is predicted to have the same direction as the \textit{obj} dependency; a leftwards arrow indicates an inverse correlation. 
	For each correlation we report what percentage of the languages in our sample satisfied it (`Real').
	We then report, for each correlation and each objective function, how many (in \%) of the optimized grammars satisfy the correlation, with the significance level in a logistic mixed-effects analysis across language families.}\label{table:results-dryer}
\end{table*}

@@

We now examine to what extent we can recover Greenberg's word order correlations in optimized grammars. 
Dryer \cite{dryer1992greenbergian} presents a comprehensive updated version of the word order correlations, drawing on 625 languages, which we take as the basis of our evaluation.
In \cite{dryer1992greenbergian}, all word order correlations are relative to the position of the direct object wrt the main verb of a sentence.
This dependency corresponds to dependencies of the form $\xrightarrow{obj}$ in UD.
Similarly, most of the other dependencies in Dryer's formulation can be formalized:
for instance, the dependency between nouns and relative clauses can be formalized as $\xrightarrow{acl}$.
Testing whether an ordering model satisfies the correlation between relative clauses and objects reduces to checking whether these two relation types have direction parameters $a_\tau$ that are either both $>0.5$ (both kinds of dependents precede their heads) or both $< 0.5$ (they both follow their heads).

@@

See the SI for details on the neural language models and parsers used, and on the optimization procedures.
}

>>>>

\begin{abstract}
The universal properties of human languages have been the subject of intense study across disciplines. We report novel computational and corpus-based evidence for the hypothesis that a prominent subset of these universal properties---those related to word order---result from a process of optimization for efficient communication among humans. We develop a probabilistic, differentiable model of word order grammars: the means by which different languages convert underlying hierarchical structures into strings of words. We show how the parameters of a word order grammar can be optimized for efficiency of information transfer, quantified as a tradeoff between incremental predictability and mutual information with latent tree structures. Applying these grammars to tree structures found in dependency corpora from 51 languages, we show that optimizing the grammar parameters for efficiency results in word order patterns that reproduce a large subset of the major word order correlations reported in the linguistic typological literature, and reproduce the predictions of previous heuristic theories such as dependency length minimization.
\end{abstract}

@@

\dropcap{F}or decades, researchers in fields ranging from philology to cognitive science to statistical physics have been involved in documenting and trying to explain the universal syntactic and statistical properties of human language \cite{behaghel1909beziehungen,zipf1949human,greenberg1963universals,lin2017critical}. An explanation for the universal properties of language would enable a deeper scientific understanding of what human language is and how to model it, with applications in psychology and natural language processing \cite{hawkins2007processing,bender2009linguistically,bender2013linguistic}. In this work we examine this question from a computational perspective, demonstrating a fully formalized framework in which certain syntactic universals can be explained through the statistical optimization of grammars for information-theoretic efficiency. % TODO add a cogsci person to citation chain at the beginning

We aim to explain the linguistic universals related to word order first documented by Greenberg \cite{greenberg1963universals}. These universals are called \key{implicational} universals, and take the form of correlations between certain word order patterns and others across languages. For example, whether a language puts verbs before or after objects strongly determines whether it places genitives before or after nouns, and whether it has prepositions or postpositions \cite{dryer1992greenbergian}.

@@

Researchers working in the functional paradigm have proposed a range of criteria that languages should meet in order to enable efficient communication and processing. These criteria are based on theories of online processing difficulty (see \cite{jaeger2011language} for a review) and on information theoretic notions of efficiency and robustness in communication \cite{ferrericancho2001two,piantadosi2011word,gibson2013noisy}, and they have been formalized to varying degrees.

@@

We formalize the notion of communicative efficiency using two terms: one having to do with minimizing the difficulty of linguistic communication, and one having to do with maximizing the information transferred in such communication. We quantify the difficulty of linguistic communication in terms of the word-by-word \key{predictability} of strings, which has been found to be a highly accurate and general predictor of human online comprehension difficulty \cite{hale2001probabilistic,levy2008expectation,smith2013effect}, and which can be justified in terms of predictive coding theories of information processing in the brain \cite{friston2009predictive} as well as the general algorithmic complexity of language generation and comprehension \cite{li2008introduction}. We quantify the information transferred in language as \key{parseability}: the quantity of information linguistic utterances contain about latent hierarchical tree structures. We also compare the results of optimizing for communicative efficiency to the results of optimizing for dependency length minimization, a previously proposed approximate metric of the amount of memory involved in language production and comprehension \cite{hudson1995measuring,gibson1998linguistic,gibson2000dependency,liu2017dependency,temperley2018minimizing}.

@@

We formalize predictability as follows. Given a language model $P_\phi$ with parameters $\phi$, the log probability of a sentence ${\bf w}$ is:
\begin{equation*}
R_{Pred}^\phi(t; {\bf w}) = \sum_{i=1}^{\#{\bf w}} \log P_\phi(w_{i+1}|w_{1\dots i}),
\end{equation*}
where the language model parameters $\phi$ are chosen to predict the language induced by tree distribution $T$ and word order grammar $\L_\theta$ as accurately as possible (maximizing predictability):
\begin{equation*}
	\phi(\theta) = \argmax_\phi \displaystyle \E_{t \sim T} \E_{{\bf w} \sim L_\theta(t)} [ R^\phi_{Pred}({\bf w}, \phi) ].
\end{equation*}
We implement the language model $P_\phi$ using an LSTM neural network architecture \cite{hochreiter1997long}, currently the state of the art for this task \cite{goldberg2017neural}. Note that the language model parameters $\phi$ and the word order grammar parameters $\theta$ are recursively dependent on each other. See the SI for details on how $\theta$ and $\phi$ are jointly optimized, as well as on the neural network architecture.

@@

\subsection{Greenbergian Word Order Correlations}
\label{sec:greenberg}
\begin{table*}
	\begin{center}
\begin{tabular}{|ll|l|l|l|ll|l|}
	\hline
	%&
	\multicolumn{2}{|c|}{Correlates with...}   &         \multirow{2}{*}{Real}   & \multirow{2}{*}{DepL}  & \multirow{2}{*}{Pred}   &  \multirow{2}{*}{Pars} &  \multirow{2}{*}{Efficiency}  \\ 
	verb & object     & & & &  & \\ \hline \hline % \textsc{verb} $\xrightarrow{obj}$ \textsc{noun}
%adp. &
adposition    &    NP       &    86    &    \textbf{81}$^{***}$    &    47    &    \textbf{76}$^{***}$    &    \textbf{68}$^{***}$   \\
copula    &    NP        &    94    &    \textbf{81}$^{***}$    &    53    &    \textbf{79}$^{***}$    &    \textbf{61}$^{**}$   \\
auxiliary    &    VP       &    88    &    \textbf{74}$^{***}$    &    \textbf{84}$^{***}$    &    55    &    \textbf{69}$^{**}$   \\
noun    &    genitive      &    80    &    \textbf{82}$^{***}$    &    55    &    \textbf{74}$^{***}$    &    \textbf{70}$^{***}$   \\
noun    &    relative clause       &    80    &    \textbf{85}$^{***}$    &    48    &    \textbf{77}$^{**}$    &    \textbf{73}$^{***}$   \\
complementizer    &    S        &    76    &    \textbf{85}$^{***}$    &    \textbf{59}$^{**}$    &    \textbf{80}$^{***}$    &    \textbf{74}$^{**}$   \\
verb    &    PP     &    88    &    \textbf{78}$^{***}$    &    \textbf{72}$^{***}$    &    59    &    \textbf{69}$^{**}$   \\
want    &    VP      &    88    &    \textbf{90}$^{***}$    &    \textbf{78}$^{**}$    &    \textbf{92}$^{***}$    &    \textbf{92}$^{***}$   \\
verb    &    subject    &    33    &    \textbf{29}$^{**}$    &    51    &    \textbf{8}$^{***}$    &    \textbf{13}$^{***}$   \\
verb    &    manner adverb     &    35    &    51    &    \textbf{21}$^{***}$    &    51    &    \textbf{32}$^{***}$   \\
 \hline
%comp. &
%\multirow{3}{*}{NP}&
%	AP &
    \multicolumn{6}{l}{\footnotesize{Significance levels: $^*$: $p < 0.05$, $^{**}$: $p < 0.01$, $^{***}$: $p < 0.001$}}
\end{tabular}
	\end{center}
	\caption{Greenbergian Correlations. Following \cite{dryer1992greenbergian}, each correlation is stated in terms of a pair of a `verb patterner' and an `object patterner', whose relative order correlate with that of verbs and objects.
%	For clarity, we organize the correlations by the category of the `verb patterner'.
	For each correlation, we give our operationalization in terms of UD. %A rightwards arrow indicates that the dependency is predicted to have the same direction as the \textit{obj} dependency; a leftwards arrow indicates an inverse correlation. 
	For each correlation we report what percentage of the languages in our sample satisfied it (`Real').
	We then report, for each correlation and each objective function, how many (in \%) of the optimized grammars satisfy the correlation, with the significance level in a logistic mixed-effects analysis across language families.}\label{table:results-dryer}
\end{table*}

@@

We now examine to what extent we can recover Greenberg's word order correlations in optimized grammars. 
Dryer \cite{dryer1992greenbergian} presents a comprehensive updated version of the word order correlations, drawing on 625 languages, which we take as the basis of our evaluation.
In \cite{dryer1992greenbergian}, all word order correlations are relative to the position of the direct object wrt the main verb of a sentence.
Most of them can be straightforwardly implemented in UD, allowing us to check which correlations a word order grammar satisfies.
%This dependency corresponds to dependencies of the form $\xrightarrow{obj}$ in UD.
%Similarly, most of the other dependencies in Dryer's formulation can be formalized:
%for instance, the dependency between nouns and relative clauses can be formalized as $\xrightarrow{acl}$.
%Testing whether an ordering model satisfies the correlation between relative clauses and objects reduces to checking whether these two relation types have direction parameters $a_\tau$ that are either both $>0.5$ (both kinds of dependents precede their heads) or both $< 0.5$ (they both follow their heads).

@@

See the SI for details on the neural language models and parsers used, and on the optimization procedures.

}

### 2019-01-25 second-draft/pnas-word-order-si.tex
attempting revision

>>>>

\section{UD Formalization}

\begin{table*}
	\begin{center}
\begin{tabular}{|ll|l|l|l|ll|l|}
	\hline
	%&
	\multicolumn{2}{|c|}{Correlates with...}   &          \multirow{2}{*}{Operationalization}       &     \multirow{2}{*}{Real}   & \multirow{2}{*}{DepL}  & \multirow{2}{*}{Pred}   &  \multirow{2}{*}{Pars} &  \multirow{2}{*}{Efficiency}  \\ 
	verb & object &      & & & &  & \\ \hline \hline % \textsc{verb} $\xrightarrow{obj}$ \textsc{noun}
%adp. &
adposition    &    NP    &    lifted\_case    &    86    &    \textbf{81}$^{***}$    &    47    &    \textbf{76}$^{***}$    &    \textbf{68}$^{***}$   \\
copula    &    NP    &    lifted\_cop    &    94    &    \textbf{81}$^{***}$    &    53    &    \textbf{79}$^{***}$    &    \textbf{61}$^{**}$   \\
auxiliary    &    VP    &    aux    &    88    &    \textbf{74}$^{***}$    &    \textbf{84}$^{***}$    &    55    &    \textbf{69}$^{**}$   \\
noun    &    genitive    &    nmod    &    80    &    \textbf{82}$^{***}$    &    55    &    \textbf{74}$^{***}$    &    \textbf{70}$^{***}$   \\
noun    &    relative clause    &    acl    &    80    &    \textbf{85}$^{***}$    &    48    &    \textbf{77}$^{**}$    &    \textbf{73}$^{***}$   \\
complementizer    &    S    &    lifted\_mark    &    76    &    \textbf{85}$^{***}$    &    \textbf{59}$^{**}$    &    \textbf{80}$^{***}$    &    \textbf{74}$^{**}$   \\
verb    &    PP    &    obl    &    88    &    \textbf{78}$^{***}$    &    \textbf{72}$^{***}$    &    59    &    \textbf{69}$^{**}$   \\
want    &    VP    &    xcomp    &    88    &    \textbf{90}$^{***}$    &    \textbf{78}$^{**}$    &    \textbf{92}$^{***}$    &    \textbf{92}$^{***}$   \\
verb    &    subject    &    nsubj    &    33    &    \textbf{29}$^{**}$    &    51    &    \textbf{8}$^{***}$    &    \textbf{13}$^{***}$   \\
verb    &    manner adverb    &    advmod    &    35    &    51    &    \textbf{21}$^{***}$    &    51    &    \textbf{32}$^{***}$   \\
 \hline
%comp. &
%\multirow{3}{*}{NP}&
%	AP &
    \multicolumn{6}{l}{\footnotesize{Significance levels: $^*$: $p < 0.05$, $^{**}$: $p < 0.01$, $^{***}$: $p < 0.001$}}
\end{tabular}
	\end{center}
	\caption{Greenbergian Correlations. Following \cite{dryer1992greenbergian}, each correlation is stated in terms of a pair of a `verb patterner' and an `object patterner', whose relative order correlate with that of verbs and objects.
%	For clarity, we organize the correlations by the category of the `verb patterner'.
	For each correlation, we give our operationalization in terms of UD. %A rightwards arrow indicates that the dependency is predicted to have the same direction as the \textit{obj} dependency; a leftwards arrow indicates an inverse correlation. 
	For each correlation we report what percentage of the languages in our sample satisfied it (`Real').
	We then report, for each correlation and each objective function, how many (in \%) of the optimized grammars satisfy the correlation, with the significance level in a logistic mixed-effects analysis across language families.}\label{table:results-dryer}
\end{table*}
